---
title: "️US Presidential Elections"
format: 
  dashboard:
    nav-buttons: [github]
logo: images/logo.png
fig-width: 10
fig-asp: 0.3
---

```{r}
#| label: load-packages
#| message: false
#| include: false
library(tidyverse)
library(DT)

```

```{r}
#| label: load-data
#| message: false
#| include: false
#| 
presidents <- read_csv("data/1976-2020-president.csv", show_col_types = FALSE)

```

#  {.sidebar}

This dashboard displays weather data for:

**Méribel, [Les Allues](https://en.wikipedia.org/wiki/Les_Allues), Auvergne-Rhône-Alpes, France**

The data were gathered from the [Virtual Crossing](https://www.visualcrossing.com/).

<br>

::: {.callout-note collapse="true"}
## Measurement units

The variables and their measurement units are given below.

|                                      |                      |
|:-------------------------------------|:---------------------|
| **Variable**                         | **Unit** |
| Temperature, Heat Index & Wind Chill | Degrees Fahrenheit   |
| Precipitation                        | Inches               |
| Snow                                 | Inches               |
| Wind & Wind Gust                     | Miles Per Hour       |
| Visibility                           | Miles                |
| Pressure                             | Millibars            |
| Solar Radiation                      | W/m^2^               |
| Solar Energy                         | MJ/m^2^              |

Source: [https://www.visualcrossing.com/resources/documentation/weather-api/unit-groups-and-measurement-units](https://www.visualcrossing.com/resources/documentation/weather-api/unit-groups-and-measurement-units/).
:::

